---
title: "AM108 Project"
author: "Ali Crump"
date: "11/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load needed libraries

library(tidyverse)
library(gt)
```

```{r}
# Read in the top 100 data and set the variable types

tennis <- read_csv("raw-data/data_top100 2.csv", col_types = cols(
                                                          Rank = col_double(),
                                                          Player = col_character(),
                                                          Points = col_double(),
                                                          Date = col_date(format = ""),
                                                          Handedness = col_character()))
```

### Baseball graph replication
I created a graph to replicate that of baseball rankings found in the paper. I calculated the percentage of lefties in tennis for each rank and plotted it. Tate later added the prediction line in R.

```{r}
# Create a column with Handedness as a dummy variable
# Add up all the lefties to find percentage of lefties in the data (by Rank)

lefty_rank <- tennis %>% 
  mutate(left = ifelse(Handedness == "Left-handed", 1,0)) %>% 
  group_by(Rank) %>% 
  summarize(perc = round(100*(sum(left)/ n())))

# Plot the percentage of lefties versus rank
# Add axis labels and remove background grid lines and gray theme

ggplot(lefty_rank, aes(x = Rank, y = perc)) + 
  geom_point(alpha = 0.8) +
  labs(x = "Rank", y = "% Left-Handed", 
       title = "Cumulative Fraction Left-handed Versus Rank", 
       subtitle = "Top 100 ATP Rankings from September 2000 to 2019") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank())

```

### Do we always see a dip?

Our original theory was that we see this dip in the top 25 rankings because of the dominance of lefty Rafael Nadal. Perhaps righties at the top were used to playing really talented lefties like Nadal, so the lefty advantage in competitive sports disappears. I filtered for years before Nadal was so dominant and used the same methods to create this graph. However, we see below that there is a "dip" in the top rankings even when Nadal isn't dominant. Therefore, there is something else going on in professional tennis that our model isn't accounting for, or this is a really odd set of circumstances that just happens by chance to be the case for the data at hand.

```{r}
# Filtering before Nadal dominant

tennis %>% 
  filter(!(Date > "2006-01-02")) %>% 
  mutate(left = ifelse(Handedness == "Left-handed", 1,0)) %>% 
  group_by(Rank) %>% 
  summarize(perc = round(100*(sum(left)/ n()))) %>% 
  filter(perc != 0) %>% 
  ggplot(aes(x = Rank, y = perc)) +
  geom_point(alpha = 0.8) +
  labs(x = "Rank", y = "% Left-Handed", 
       title = "Cumulative Fraction Left-handed Versus Rank", 
       subtitle = "Top 100 ATP Rankings from September 2000 to 2005") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank())
```

### Data exploration

We were curious to see how many lefties had even been in the top 25 rankings consistently, or if it really was just Nadal. We find that 6 left handers have consistently been in the top 25, consistently meaning in the top 25 for more than a year. We were suprised by how low this number was.

```{r}
top <- tennis %>% 
  filter(Rank < 25) %>% 
  group_by(Player) %>% 
  mutate(n = n()) %>% 
  filter(n > 52)

top %>%
  filter(Handedness == "Left-handed") %>% 
  distinct(Player, keep.all = TRUE) %>% 
  select(Player)
  
```

### Left-handedness over time

We were curious as to how the left handed population changes over time, but it's very messy because I didn't know how to plot it by year, so the graph shows the lefty percent for 52 weeks of rankings for 20 years.

```{r}
tennis %>% 
  mutate(left = ifelse(Handedness == "Left-handed", 1,0)) %>% 
  group_by(Date) %>% 
  mutate(perc = sum(left)/100) %>% 
  ggplot(aes(x = Date, y = perc)) +
  geom_jitter() +
  labs(y = "% Left-handed")
```

### Agent-based model results 

We used an agent-based model to simulate the left handed population in tennis for many years.  

```{r}
# There is a much nicer way of reading in all of this data, but we were limited
# with our time. Here I'm reading in the data for each simulation, and I'm 
# naming the columns and specifying the variable types. I repeat this 12 times
# and then combine all of the data into one data frame.

# lefty percetage: 0.0859375
simulation0 <- read_csv("raw-data/data.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))
# 0.082
simulation1 <- read_csv("raw-data/data1.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))
#  0.1055
simulation2 <- read_csv("raw-data/data2.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))

#  0.1055
simulation3 <- read_csv("raw-data/data3.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))
# 0.09765625
simulation4 <-  read_csv("raw-data/data4.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))
# 0.12109375
simulation5 <-  read_csv("raw-data/data5.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))
# 0.09375
simulation6 <-  read_csv("raw-data/data6.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))
# 0.0703125
simulation7 <-  read_csv("raw-data/data7.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))

# 0.078125
simulation8 <-  read_csv("raw-data/data8.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))

# 0.08984375
simulation9 <-  read_csv("raw-data/data9.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))
# 0.09375
simulation10 <-  read_csv("raw-data/data10.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))
# 0.13671875
simulation11 <-  read_csv("raw-data/data11.csv", col_names = c("rank", "percent"), 
                       col_types = cols(
                                        rank = col_double(),
                                        percent = col_double()))

simulations <- rbind(simulation0,simulation1, simulation2, simulation3, simulation4, simulation5, simulation6, simulation7, simulation8, simulation9,simulation10, simulation11)

```

This graph shows the results of one simulation where we see a dip, like what we saw with our actual tennis data.


```{r}
# Simple graph of our results from one simulation. We're graphing percent of
# left handers for each rank.

simulation0 %>% 
  mutate(percent = 100*percent) %>% 
  ggplot(aes(x = rank, y = percent)) +
  geom_point() +
  labs(x = "Rank", y = "% Left-Handed", 
       title = "One Simulation of Our Agent-Based Model") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank())
```

This graph then shows the average of all of our simulations, we no longer see a dip, as the dip only happened about 30% of the time in our simulations.


```{r}
# We're taking the average of % lefties for each ranking respectively.
# Then we're making a scatter plot of our results, similar to our 
# previous graphs. I set the ylimits to be the same as the previous
# graph so that we could compare the two more easily.

simulations %>%
  group_by(rank) %>% 
  mutate(avg = 100*mean(percent)) %>% 
  ggplot(aes(x = rank, y = avg)) +
  geom_point() +
  labs(x = "Rank", y = "Avg % Left-Handed", 
       title = "Twelve Simulations of Our Agent-Based Model") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
  ylim(0.0,50)
```
